---
title: "Sample size for the verification of possible Interleukin-6 circadian rhythm in Alpha-1 Antitrypsin deficient patients"
subtitle: 'DOCUMENT: SAP-2021-020-LH-v01'
author: '**From:** Felipe Figueiredo **To:** lameesehadda (fiverr.com)'
# date: yyyy-mm-dd
output:
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    keep_md: yes
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: yes
    toc: yes
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    reference_docx: misc/style_SAP_en.docx
    toc: yes
# toc-title: "Sumário"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
options(scipen = 999)
library(pander)
library(knitr)
library(philsfmisc)
panderOptions('table.style', 'rmarkdown')
panderOptions('table.split.table', 160)
```

---

**Sample size for the verification of possible Interleukin-6 circadian rhythm in Alpha-1 Antitrypsin deficient patients**

**Document version**

```{r history}
Version <- c("01")
Changes <- c("Initial version")
history <- cbind(Version, Changes)
colnames(history) <- c("Version", "Alterations")
# pander(history, split.cells = 70)
kable(history)
```

---

# Abbreviations

- BMI: body mass index
- CI: confidence interval
- d: Cohen's D effect size
- IQR: interquartile range
- SD: standard deviation

# Context

## Objectives

Determine the sample size to detect the difference of the daily range of Interleukin-6 blood levels between Alpha-1 Antitrypsin deficient and healthy patients.

## Hypotheses

The circadian rhythm in patients induces a detectable amplitude between peak and lower daily IL6 levels.
The average IL6 daily amplitudes differ between Alpha-1 Antitrypsin deficient and healthy patients.

# Data

```{r cleaning, include=FALSE}
source("scripts/input.R", encoding = 'UTF-8')
```

## Raw data

A unique study ID will be created for each participant to allow tracking of consecutive IL6 blood level measurements.
IL6 data of each participant will be collected at three different times per day.
Each datum will have its respective date and time of speciment collection recorded.
Participant demographic characteristics to be collected are the group to with they belong (AAT-defficient or healthy), the date of birth, sex and BMI at the date of inclusion.

After the cleaning process the raw data will include `r ncol(raw_mockup)` variables.
Table 1 shows the structure of the raw dataset.

```{r analytical data}
raw_mockup %>%
  pander(caption = "**Table 1** Raw dataset structure.")
```

## Analytical dataset

Age at inclusion will be computed between the date of birth and the date of the first specimen collection.

All variables in the analytical set will be labeled according to the raw data provided and values were labeled according to the data dictionary for the preparation of production-quality results tables and figures.

The analytical dataset will be included in the private version of the report, and will be omitted from the public version of the report.

# Study variables

## Primary and secondary outcomes

## Covariates

# Statistical methods

## Statistical analyses

### Descriptive analyses

N/A

<!-- The epidemiological profile of the study participants will be described. -->
<!-- Demographic -->
<!-- (sex, age and BMI) -->
<!-- and clinical variables -->
<!-- ( [vars] ) -->
<!-- will be described as -->
<!-- mean (SD) -->
<!-- median (IQR) -->
<!-- or as counts and proportions (%), as appropriate. -->
<!-- The distributions of participants' characteristics will be summarized in tables and visualized in exploratory plots. -->

### Inferential analyses

N/A

<!-- All comparisons between groups will be performed as univariate analyses. -->
<!-- Continuous variables will be compared between groups with the -->
<!-- independent t test with Welch correction. -->
<!-- paired t test with Welch correction. -->
<!-- Wilcoxon test. -->
<!-- Differences in distribution of categorical variables will be assessed with the -->
<!-- Fisher exact test. -->
<!-- chi-square test. -->
<!-- McNemar test. -->

<!-- All inferential analyses will be performed in the statistical models (described in the next section). -->

### Statistical modeling

N/A

## Significance and Confidence Intervals

All analyses will be performed using the significance level of 5%.
All significance hypothesis tests and confidence intervals computed will be
two-tailed.
<!-- left-tailed. -->
<!-- right-tailed. -->

## Study size and Power

```{r}
n <- analytical %>%
  count(group) %>%
  pull(n)
n1 <- n[1]
n2 <- n[2]
total <- sum(n)
d <- pwr::pwr.t2n.test(n1 = n1, n2 = n2, power = .8, sig.level = .05)$d %>%
  format.float()
d <- pwr::pwr.t.test(n = total/2, power = .8, sig.level = .05)$d %>%
  format.float()
```

<!-- Study groups were balanced and the total sample size was `r total` including all groups. -->

<!-- Study groups were not balanced with -->
<!-- group1 with `r n1` participants -->
<!-- and -->
<!-- group2 with `r n2` participants -->
<!-- with a total study size of `r total`. -->
<!-- This power analysis adjusts for different group sizes. -->

<!-- With these group sizes a t test can detect an effect size as large as -->
<!-- **d = `r d`** -->
<!-- with 80% power and 5% significance level (Cohen, 1988). -->

N/A

## Statistical packages

This analysis will be performed using statistical software `R` version `r getRversion()`.

# Observations and limitations

**Direction of the difference**

The AAT deficient patient IL6 blood levels may be hypothesized to be lower (higher) than the levels observed healthy patients.
If this hypothesis holds, a left tailed (right tailed) test will have more statistical power to detect significant differences than a two-tailed test.
If there is support in the literature for an educated guess on the direction of the difference this could improve the chance of having significant p-values in the final analysis, at the risk of not detecting a difference in the other direction.

For example, if the hypothesis is specified in the left direction (group of interest has lower levels of IL6 than the healthy group) and the actual observed levels indicate that the group of interest has higher levels of inflammatory biomarkers, then the actual difference will likely not be significant in the study.
It is recommended that such decision be made with caution.

# References

- **SAR-2021-020-LH-v01** -- Sample size for the verification of possible Interleukin-6 circadian rhythm in Alpha-1 Antitrypsin deficient patients
<!-- - Cohen, J. (1988). Statistical power analysis for the behavioral sciences (2nd Ed.). New York: Routledge. -->

# Appendix

## Availability

Both this analytical plan and the corresponding analysis report (**SAR-2021-020-LH-v01**) can be downloaded in the following address:

<https://philsf-biostat.github.io/SAR-2021-020-LH/>
